What should I notice? Evaluating the memorability of events for abductive inference.

### 2021-10-14 new_paper.tex
expanding results and related

\begin{figure}[ht]
    \centering
    \includegraphics[width=\linewidth]{figures/simulator}
    \caption{View of the simulator's web interface provided by iCasa. The four
        rooms are visible, with their equipment and the user.}
    \label{fig:view}
\end{figure}

@@

The values observed from all devices and zones was sampled throughout
the simulation run. The resulting data (figure
\ref{fig:ts_example}) was then used as a basis for our experiments.

@@

For the implementation of our method, we first needed to identify and
characterize events from the time series data generated by the iCasa simulation.
Since the selection of events is not the focus point of our present work (see sec.
\ref{sec:related}), we perform event detection merely based on threshold and pre-computed conditions.

This set of events constitutes the basis of the initial memory $M_0$ used
for computations.
Then we implemented a small number of predicate concepts:
\begin{itemize}
    \item $\mathtt{label(e, k)}$: whether the event $e$ has the label ranked
          $k^{th}$ in the label list.
    \item $\mathtt{rank(e, r, a)}$: whether the event $e$ has the rank $r$
          along axis $a$.
    \item $\mathtt{day(e, k)}$: whether the event $e$ occurred $k$ days ago.
    \item $\mathtt{month(e, k)}$: whether the event $e$ occurred $k$ months
          ago.
    \item $\mathtt{location(e, k)}$: whether the event $e$ occurred in zone
          $k$.
\end{itemize}

@@

\begin{figure}
    \centering
    \includegraphics[width=.8\linewidth]{./figures/abduction_1_results}
    \caption{Outputs of the memorability-based abduction module. This represent the 3 events in the memory having the highest memorability score relative to the consequence event.}
    \label{fig:abduction_res}
\end{figure}
\subsubsection{The ``temperature'' scenario:}
\paragraph{Memorability}

@@

The blue line emerges from a set of ``usual'' events, whose complexity varies as the logarithm of
the elapsed time since their occurrence. It corresponds to events
for which the best retrieval path consists of a time description (e.g. ``2
months and 12 days ago''). On the other hand, some events stand out in terms of
complexity: some appear simpler, as they can be distinguished by using their
rank along some axis (``the hottest day'', ``the second longest user's
absence''), or the rare occurrence of their kind (``the only fault on the heater'').

@@

The ``memorability score'' is shown in Figure \ref{fig:result1}.
% highlights these aforementioned events from the main ``usual'' sequence. Since
the computation of this measure treats unusually complex or unusually simple events the
same way (from the absolute value operation in Equation \ref{eq:unexpected}), we
observe that some events are memorable due to their context only. For instance,
a temperature anomaly occurring simultaneously to many other anomalies is
noticeable, as it is costlier than expected to distinguish it from its neighbors.

@@

@@

\begin{figure}[ht]
    \centering
\includegraphics[width=0.7\linewidth]{./figures/roc}
    \caption{Experimental ROC curve (True Positive Rate against False Positive
Rate) for a classifier based on our memorability score. Measures consider 23
manually flagged events as memorable (events added to the normal background
as described in Section~\ref{sec:example}.)}
    \label{fig:roc}
\end{figure}
\subsubsection{Abduction}

@@

\textbf{TODO: the experiment still needs to be done: some additional data is required here (creating the scenario and finding causes.)}

@@

\subsection{Related Works}
\label{sec:related}
\textbf{TODO: this section is still WIP: I've added most references and short comments, but I still need to write the story linking them.}
Our work is intended to be integrated into larger-scale frameworks to monitor
and detect events in complex environment such as smart homes. In these works, the
approach to smart homes is often regarded as self-organizing systems \cite{kramer_rigorous_2009,kounev_notion_2017}. As such, they present capacities of adaptation to new goals,
new components, new environment. A commonly used approach is the principle of
autonomic system, which minimizes user's intervention for management of the
system \cite{kounev_notion_2017,kephart_vision_2003}.

@@

More recently, \cite{marx_causal_2018} used Minimum Description Lenght to determine, given a joint probability distribution over $(X,Y)$, whether $X$ causes $Y$ or $Y$ causes $X$. Their method is based on tree models, and implying that a model respecting the causal relation will be simpler to describe.

\cite{tatti_finding_2008} PACK algorithm to cluster data using the simplest possible trees: maybe link this to isolation forests?

The topic of finding events from streams of time series data has already been explored in many ways.
\cite{aggarwal_outlier_2017} provides good review of modern approaches and techniques
in the field. Some previous work can also be noted for having used AIT techniques to qualify and detect events in time series data. For instance, \cite{batista_complexity-invariant_2011,fadlallah_weighted-permutation_2013} propose weighted permutation entropy as a proxy for complexity measures in time series data, and use it to find relations between different time series. \cite{hu_discovering_2011} proposes a MDL approach to find the intrinsic dimensions of time series. All these approaches are interesting, and can be integrated into our canvas as tools to detect events using only complexity. As such, one could acheive a purely complexity-driven process for detecting and qualifying events.

Among various methods, the approach of ``Isolation forests''\cite{liu_isolation_2008,hariri_extended_2021} is closely related, by design, to ours. It evaluates the isolation of data points by constructing random binary tree classifiers. On average, outlier points will require less operations to be singled out. Using the average height of leaves in the tree as a metrics, this approach succeeds in identifying outlier points without having to define a ``typical'' point. This approach can be understood in terms of complexity: each node of binary tree classifier needs a fixed amount of information to be described (which variable and threshold are used). So nodes that place higher need less information to be described. As such, outliers need less information to be singled out. Compared to ours, this method is tailored for data-points living in the same metric space. By using predicates as a proxy for complexity computation, our methods is more general, as it is agnostic regarding the nature of events. However, the introdcution of predicates adds a subjectivity in the determination of memorable points, as we will discuss later.

>>>>

\begin{figure}[ht]
    \centering
\includegraphics[width=.8\linewidth]{figures/simulator}
    \caption{View of the simulator's web interface provided by iCasa. The four
        rooms are visible, with their equipment and the user.}
    \label{fig:view}
\end{figure}

@@

The values observed from all devices and zones was sampled throughout
the simulation run. The resulting data (figure
\ref{fig:ts_example}) was then used as a basis for our experiments.
We then process the time series data to identify and characterize events.
Since the selection of events is not the focus point of our present work (see sec.
\ref{sec:related}), we perform event detection merely based on threshold and pre-computed conditions.

@@

We implemented the computation of both the descritpion complexity and the memorability score into a Python object, called the \texttt{SurpriseAbductionModule}. Apart from the base memory of events $M_0$, this module contains a set of predifined predicates $\Pi$ to characterize events. For instance, for scenario 2, the predicates we use are the following:
\begin{itemize}
    \item $\mathtt{label(e, k)}$: whether the event $e$ has the label ranked
          $k^{th}$ in the label list.
    \item $\mathtt{rank(e, r, a)}$: whether the event $e$ has the rank $r$
          along axis $a$.
    \item $\mathtt{day(e, k)}$: whether the event $e$ occurred $k$ days ago.
    \item $\mathtt{month(e, k)}$: whether the event $e$ occurred $k$ months
          ago.
    \item $\mathtt{location(e, k)}$: whether the event $e$ occurred in zone
          $k$.
\end{itemize}

The description length $L_{\pi, k}$ of using a predicate is computed as follows: since the set of predicates is finite and known, $L(\pi) = \log2(|\Pi|)$ are enough to describe the predicate concept $\pi$\footnote{This approach gives an equal complexity to all predicate concepts. While this could be argued when using many concepts, as humans do, we deemed better, for our small examples, to consider all concepts as equal.}. To encode the argument $k$ of the predicate, we used the prefix-free Elias delta code, which requires $L(k) = \log2(k) + 2 \log2(\log2(k)+1) + 1]$ bits. The total cost of decribing $\pi_{k}$ therefore is
\begin{equation}
    \L(\pi, k) = \log2(|\Pi|) + \log2(k) + 2 \log2(\log2(k) + 1) + 1
\end{equation}

@@

\begin{figure}
    \centering
    \includegraphics[width=.8\linewidth]{./figures/abduction_1_results}
    \caption{Outputs of the memorability-based abduction module. This represent the 3 events in the memory having the highest memorability score relative to the consequence event.}
    \label{fig:abduction_res}
\end{figure}

\subsubsection{The ``temperature'' scenario:}
\paragraph{Memorability}

@@

The blue line emerges from a set of ``usual'' events, whose complexity varies as the logarithm of
the elapsed time since their occurrence. It corresponds to events
for which the best retrieval path consists of a time description (e.g. ``2
months and 12 days ago''). As such, it appears that most days are considered usual by our memorability score: therie is no better way to retrieve them than simply giving the day of their occurence. On the other hand, some events stand out in terms of
complexity: some appear simpler, as they can be distinguished by using their
rank along some axis (``the hottest day'', ``the second longest user's
absence''), or the rare occurrence of their kind (``the only fault on the heater'').

@@

The ``memorability score'' is shown in Figure \ref{fig:result1}. Similar to what happened with the description complexity score, most events appear with a low memorability: this corresponds mostly to events from the ``main sequence'' from Figure. \ref{fig:computed_cplx}. On the other hand, some events stand out : for instance event \textbf{TODO}, which are the hottest and coldest days recorded, or event \textbf{TODO} which correspond to the rare type \emph{device\_removal}.
Since our memorability measure treats unusually complex or unusually simple events the
same way (from the absolute value operation in Equation \ref{eq:unexpected}), we
observe that some events are memorable due to their context only. For instance,
the group \textbf{TODO} appears more complex than expected: the same event occuring simultaneously in all four rooms of the house make each instance harder to discern. Table \ref{tab:predicates}.

@@

\begin{table}
    \centering
    \begin{tabular}{c|c|c}
        Event iD & Event Type    & Retrieval Path                         \\
        \hline
        1        & hot           & Label("hot"), AxisRank(1)              \\
        2        & hot           & Label("hot"), Day(21), Location(room2) \\
        102      & deviceRemoval & Label("deviceRemoval), Day(2)          \\
    \end{tabular}
    \caption{Selected events from the ``temperature example'', with their shortest retrieval path.}
\end{table}

@@

\begin{figure}[ht]
    \centering
\includegraphics[width=0.7\linewidth]{./figures/roc}
    \caption{Experimental ROC curve (True Positive Rate against False Positive
Rate) for a classifier based on our memorability score. Measures consider 23
manually flagged events as memorable (events added to the normal background
as described in Section~\ref{sec:example}.)}
    \label{fig:roc}
\end{figure}

@@

\textbf{TODO: the figure is still missing!}

@@

\subsection{Related Works}
\label{sec:related}
Our work is intended to be integrated into larger-scale frameworks to monitor
and detect events in complex environment such as smart homes. In these works, the
approach to smart homes is often regarded as self-organizing systems \cite{kramer_rigorous_2009,kounev_notion_2017}. As such, they present capacities of adaptation to new goals,
new components, new environment. A commonly used approach is the principle of
autonomic system, which minimizes user's intervention for management of the
system \cite{kounev_notion_2017,kephart_vision_2003}.

@@

The relation between complexity, compression and causality was used in \cite{tatti_finding_2008} to devise the PACK algorithm. It models a dataset by using a family of decision trees where each tree describes how one variable can be expressed given the others. By chosing the model minimizing the total description length (i.e. description of the model and description of the errors), PACK compresses the dataset while finding relations between variables which can be further analyzed. More recently, \cite{marx_causal_2018} used Minimum Description Lenght to determine, given a joint probability distribution over $(X,Y)$, whether $X$ causes $Y$ or $Y$ causes $X$. Their method is based on tree models, and implying that a model respecting the causal relation will be simpler to describe.

Another topic where AIT can provide original approaches is event mining in data streams.
\cite{aggarwal_outlier_2017} provides a good review of modern approaches and techniques
in the field. Some previous work can also be noted for having used AIT techniques to qualify and detect events in time series data. For instance, \cite{batista_complexity-invariant_2011,fadlallah_weighted-permutation_2013} propose weighted permutation entropy as a proxy for complexity measures in time series data, and use it to find relations between different time series. \cite{hu_discovering_2011} proposes a MDL approach to find the intrinsic dimensions of time series. All these approaches are interesting, and can be integrated into our canvas as tools to detect events using only complexity. Using this, one could acheive a purely complexity-driven process for detecting and qualifying events.

The philosphy of our approach can be closely related to the ``Isolation forests'' method\cite{liu_isolation_2008,hariri_extended_2021}. It evaluates the isolation of data points by constructing random binary tree classifiers. On average, outlier points will require less operations to be singled out. Using the average height of leaves in the tree as a metrics, this approach succeeds in identifying outlier points without having to define a ``typical'' point. This approach can be understood in terms of complexity: each node of binary tree classifier needs a fixed amount of information to be described (which variable and threshold are used). So nodes that place higher need less information to be described. As such, outliers need less information to be singled out. Compared to ours, this method is tailored for data-points living in the same metric space. By using predicates as a proxy for complexity computation, our methods is more general, as it is agnostic regarding the nature of events. However, the introdcution of predicates adds a subjectivity in the determination of memorable points, as we will discuss later.
